### 2012-10-16 paper_sean/abstract.tex
JBY edits and comments.

% !TEX root =main.tex
Creating gaits for physical robots is a longstanding and still open challenge. 
%Evolutionary algorithms can outperform human engineers in generating such gaits, but this approach historically required engineers to manually decide which types of gaits would evolve. Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Recently, the HyperNEAT generative encoding was shown to automatically discover a variety of gait regularities, producing fast, coordinated gaits, but only for simulated robots. A follow-up study found that HyperNEAT did not produce impressive gaits when they were evolved directly on a physical robot. A simpler encoding hand-tuned to produce regular gaits was tried on the same robot, and outperformed HyperNEAT, but these gaits were first evolved in simulation before being transferred to the robot. In this paper, we tested the hypothesis that the beneficial properties of HyperNEAT would outperform the simpler encoding if HyperNEAT gaits are first evolved in simulation before being transferred to reality. 
That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing techniques for generating gaits for this robot. This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, were not limited to evolving gaits in simulation, but indeed work on real robots.

>>>>

% !TEX root =main.tex
Creating gaits for physical robots is a longstanding and open challenge. 
%Evolutionary algorithms can outperform human engineers in generating such gaits, but this approach historically required engineers to manually decide which types of gaits would evolve. Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Recently, the HyperNEAT generative encoding was shown to discover automatically a variety of gait regularities, producing fast, coordinated gaits, but only for simulated robots. A follow-up study found that HyperNEAT did not produce impressive gaits when they were evolved directly on a physical robot. A simpler encoding hand-tuned to produce regular gaits was tried on the same robot, and outperformed HyperNEAT, but these gaits were first evolved in simulation before being transferred to the robot. In this paper, we tested the hypothesis that the beneficial properties of HyperNEAT would outperform the simpler encoding if HyperNEAT gaits are first evolved in simulation before being transferred to reality. 
That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing techniques for generating gaits for this robot. This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, were not limited to evolving gaits in simulation, but indeed work on real robots.

### 2012-10-16 paper_sean/conclusion.tex
JBY edits and comments.

%Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
%When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
However, we faced a problem in the transfer from simulation to reality. %because of the physical limitations of the robot, these exploitations did not transfer well to reality. %could not be transferred well to reality.
One of the reasons was that the robot was greatly underpowered and could not perform the transferred gaits. The other reasons was the inevitable faults with the simulator (as it cannot model the robot or the real world perfectly). Because of these factors, many gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.

Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality by reducing the \emph{reality gap}~\cite{jakobi}. Many different approaches to reduce the reality gap has been proposed~\cite{koos2010crossing,bongard,zagal}. One of the easiest and most promising of these methods is the Transferability Approach~\cite{koos2011transferability}, which has already proven to be successful in quadrapedal gait evolution. Another method that we could take is to evolve the simulator itself to adapt itself to the real world. HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait for many robotic platforms.

>>>>

%Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
%When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
However, we faced a problem in the transfer from simulation to reality. %because of the physical limitations of the robot, these exploitations did not transfer well to reality. %could not be transferred well to reality.
One of the reasons was that the robot was greatly underpowered and could not perform the transferred gaits. The other reasons was the inevitable faults with the simulator (as it cannot model the robot or the real world perfectly). Because of these factors, many gaits performed much worse than they did in simulation and had difficulty in achieving consistent results.

Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality by reducing or managing the \emph{reality gap}~\cite{jakobi}. Many different approaches to reduce the reality gap have been proposed~\cite{koos2010crossing,bongard,zagal}. One of the easiest and most promising of these methods is the Transferability Approach~\cite{koos2011transferability}, which has already proven to be successful in quadrapedal gait evolution. Another method that we could take is to evolve the simulator itself to adapt itself to the real world. HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait for many robotic platforms.

### 2012-10-16 paper_sean/hyperneat.tex
JBY edits and comments.

% !TEX root =main.tex
HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2009hypercube}. It has been repeatedly described in detail~\cite{stanley2009hypercube,gauci2007generating,clune2011performance},  so here we provide only a summary. Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007CPPN}. The CPPN genome specifies the weights in a similar way to how natural organisms develop. In nature, phenotypic attributes are specified as a function of their geometric location, and such positional information is conveyed through chemical morphogen gradients~\cite{carroll2005endless}. For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd-numbered or even-numbered segment. Based on the relative concentrations of these chemicals, a cell can know where it is geometrically and, thus, what type of cell to become~\cite{carroll2005endless}.

With CPPNs this process is abstracted as a network of math functions that operate in a Cartesian geometric space. The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait. For example, when CPPNs encode 2D pictures, the coordinates of each pixel are iteratively input into the genome and the output is the grayscale value at that coordinate~\cite{secretan2011picbreeder}. Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype. For example, a Gaussian function of the x-axis can provide left-right symmetry, and a repeating function (e.g. sine) of the y-axis could provide a related motif such as segmentation~(\figref{hyperneat_bug_example}). Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.

@@

In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al. 2011~\cite{yos:clune}. 
The ANN has a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
The inputs to the substrate were the angles requested in the prevous time step for each of the 9 joints of the robot and a sine and cosine wave to facilitate periodic motion. 
The outputs of the substrate at each time step were nine numbers (for each joint) in the range [-1, 1] which were scaled to the range [0, 1023], the allowable ranges for the servos. 
As in Yosinski et al. 2011~\cite{yos:clune}, to obtain each commanded position we averaged over four outputs of the network to reduce the possibility of switching from extreme negative to extreme positive numbers at a very high frequency.

>>>>

% !TEX root =main.tex

\subsubsection{HyperNEAT.}

\editbox{JBY: this whole section could be shortened if needed. At the very least, it could be rewritten to seem less like it was ripped out of another paper. Work on this last.}

HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2009hypercube}. It has been repeatedly described in detail~\cite{stanley2009hypercube,gauci2007generating,clune2011performance},  so here we provide only a summary. Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007CPPN}. The CPPN genome specifies the weights in a similar way to how natural organisms develop. In nature, phenotypic attributes are specified as a function of their geometric location, and such positional information is conveyed through chemical morphogen gradients~\cite{carroll2005endless}. For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd-numbered or even-numbered segment. Based on the relative concentrations of these chemicals, a cell can know where it is geometrically and, thus, what type of cell to become~\cite{carroll2005endless}.

With CPPNs this process is abstracted as a network of mathematical functions that operate in a Cartesian geometric space. The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait. For example, when CPPNs encode 2D pictures, the coordinates of each pixel are iteratively input into the genome and the output is the grayscale value at that coordinate~\cite{secretan2011picbreeder}. Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype. For example, a Gaussian function of the x-axis can provide left-right symmetry, and a repeating function (e.g. sine) of the y-axis could provide a related motif such as segmentation~(\figref{hyperneat_bug_example}). Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.

@@

In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al. 2011~\cite{yos:clune}. 
The ANN has a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
The inputs to the substrate were the angles requested in the prevous time step for each of the 9 joints of the robot and a sine and cosine wave to facilitate periodic motion. 
The outputs of the substrate at each time step were nine numbers (for each joint) in the range [-1, 1] which were scaled to the range [0, 1023], the allowable ranges for the servos. 
As in Yosinski et al. 2011~\cite{yos:clune}, we generated pseudo-positions at 160Hz and then downsampled over consecutive blocks of 4 to obtain the actual commanded positions at 40Hz; this reduced the number of gaits which commanded switches from extreme negative to extreme positive numbers at 40Hz, an undesirable artifact which only served to tax the servos.

### 2012-10-16 paper_sean/methods.tex
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist
JBY edits and comments.

% !TEX root =main.tex
We evolved gaits first in simulation and then transferred them to the physical QuadraTot robot, as in Glette et al. \cite{glette}. 
Each run had a population size of 200 and lasted 200 generations. 20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation. 
Each gait was evaluated for 12 seconds, both in simulation and in reality. 
All HyperNEAT parameters were identical to those in Yosinski et al. 2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns do to high-frequency gaits.

@@

 %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot. 
If the average frequency of each leg for the gait was higher than the threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:

\begin{equation}
distance = distance*e^{(freq_{thr}-freq)} \\
\end{equation}

\noindent where \emph{d} is the Euclidean distance the center of mass of the robot travelled during the gait duration. Following Clune et al. 2009~\cite{clune2009evolving}, the fitness equation was

\begin{equation}
fitness = 2^{distance^{2}}
\end{equation}

>>>>

% !TEX root =main.tex

\subsection{Methods}

We evolved gaits first in simulation and then transferred them to the physical QuadraTot robot, as in Glette et al. 2012~\cite{glette}. 
Each run had a population size of 200 and lasted 200 generations. 20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation. 
Each gait was evaluated for 12 seconds \edit{JBY: how? As in fig 4 of \cite{yos:clune}, or another way?}, both in simulation and in reality. 
All HyperNEAT parameters were identical to those in Yosinski et al. 2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns due to high-frequency gaits \edit{JBY: what? why?}.

@@

 %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot. 
If the average frequency of each leg \edit{JBY: any leg or all legs?} for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:

\begin{equation}
\mathrm{distance} = \mathrm{distance} \cdot e^{(\mathrm{freq}_{\mathrm{thr}}-\mathrm{freq})} \\
\end{equation}

\noindent where \emph{$\mathrm{distance}$} is the Euclidean distance the center of mass of the robot travelled during the gait duration. Following Clune et al. 2009~\cite{clune2009evolving}, the fitness equation was

\begin{equation}
\mathrm{fitness} = 2^{\mathrm{distance}^{2}}
\end{equation}

### 2012-10-16 paper_sean/results.tex
JBY edits and comments.

% !TEX root =main.tex
Because of the noise inherent in all physical robots, especially with different copies of robots, the best way to compare two algorithms is in simulation. The gaits evolved with HyperNEAT in simulation significantly outperformed the gaits evolved with a GA by Glette et al. 2012~(Table~\ref{resultsTable}). We compared Kyrre et al 2012's best simulated run to the average of our 20 runs and found that our gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s standard deviation, for every generation of evolution during the runs. Using the same data, we recorded a fitness of 21.0 cm/s with 4.29 cm/s standard deviation as the mean fitness throughout the evolutionary process while Kyrre et al 2012's data recorded a fitness of 13.35 cm/s with 4.06 cm/s standard deviation. These results show that in simulation, HyperNEAT's gaits evolve faster and have higher fitnesses than those of GA's.

@@

HyperNEAT+Simulator gaits also outperformed the best gait evolved with HyperNEAT directly in hardware~\cite{yos:clune} by 49.5\%, most likely due to the difference in the population size (200 vs. 9) and the number of generations (200 vs. 20), leading to a total difference in the number of evaluations between the two studies of 40000 vs 180~\cite{yos:clune}. Another potential cause of improved performance is that with a simulator HyperNEAT had much less noise in the evaluation process during early generations: it could thus find coordinated, regular, gaits, which are both higher-performing and more likely to transfer. On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}. Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty~(\figref{avg_fit_vs_gen}). It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.

% table!
\begin{table}
\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s]. Data used from Yosinksi et al. 2011~\cite{yos:clune}, Glette et al. 2012~\cite{glette}, and Shen et al. 2012~\cite{haocheng}.
*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline
                                         & Evaluations  & Simulated Velocity  & Real Velocity \\
\hline
Parameterized gaits + optimization    &153    & --    & 4.6 \\
\hline
HyperNEAT in hardware                  & 180         & --         &   9.7     \\
\hline
RL PoWER Spline                          & 300         & --         &   11.1 \\
\hline
GA + simulator              & 60000       & 16.4       &   *13.0     \\
\hline
HyperNEAT + simulator                      & 40000       & 25.4       &   14.5 \\
\hline
\end{tabular}
\end{center}
\label{resultsTable}
\end{table}

Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving,clune2011performance}. This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~(\figref{servo_plot_111}). This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation. This conclusion was not obvious after HyperNEAT failed to produce substantially coordinated, regular, effective gaits on this robot platform when evolving directly on the hardware~\cite{yos:clune}.

@@

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=fit_gen.eps, width=8cm}
% \caption[ ]{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s]. These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}
% \end{center}
% \end{figure}
\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{Performance plots. \textbf{Left: }Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s]. These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%. \textbf{Right: }Average frequencies of the gaits averaged over 20 runs. After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits. This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.}

@@

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=servo_pos.eps, width=8cm}
% \caption[ ]{Top: A plot of servo positions for a gait produced in this study. Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware. Figure taken from Yosinski et al.~\cite{yos:clune}. The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
% \end{center}
% \end{figure}
%\figgp{servo_plot_111}{.6}{neat_110115_211410_00000_002_filt_zoom}{.6}{Top: A plot of servo positions for a gait produced in this study. Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware. Figure taken from Yosinski et al.~\cite{yos:clune}. The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}
\figp{servo_plot_111}{0.95}{A plot of servo positions for a gait produced in this study. The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}

>>>>

% !TEX root =main.tex

\edit{JBY: IMHO, this section could use the most work. Maybe we can attack it together, Sean?}

Because of the noise inherent in all physical robots, especially with different copies of robots, the best way to compare two algorithms is in simulation. The gaits evolved with HyperNEAT in simulation significantly outperformed the gaits evolved with a GA by Glette et al. 2012~(Table~\ref{resultsTable}). We compared Kyrre et al 2012's best simulated run to the average of our 20 runs and found that our gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s standard deviation, for every generation of evolution during the runs. Using the same data, we recorded a fitness of 21.0 cm/s with 4.29 cm/s standard deviation as the mean fitness throughout the evolutionary process while Kyrre et al 2012's data recorded a fitness of 13.35 cm/s with 4.06 cm/s standard deviation. These results show that in simulation, HyperNEAT's gaits evolve faster and have higher fitnesses than those of GA's.

@@

HyperNEAT+Simulator gaits also outperformed the best gait evolved with HyperNEAT directly in hardware~\cite{yos:clune} by 49.5\%, most likely due to the difference in the population size (200 vs. 9) and the number of generations (200 vs. 20), leading to a total difference in the number of evaluations between the two studies of 40000 vs 180~\cite{yos:clune}. Another potential cause of improved performance is that with a simulator HyperNEAT had much less noise in the evaluation process during early generations: it could thus find coordinated, regular, gaits, which are both higher-performing and more likely to transfer. On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}. Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty~(\figref{avg_fit_vs_gen}, right). It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.

% table!
\begin{table}
\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s]. Data used from Yosinksi et al. 2011~\cite{yos:clune}, Glette et al. 2012~\cite{glette}, and Shen et al. 2012~\cite{haocheng}.
*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s.}  %\tablabel{results}
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline
                                         & Evaluations  & Simulated Velocity  & Real Velocity \\
\hline
Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 4.6 \\
\hline
HyperNEAT in hardware \cite{yos:clune}                 & 180         & --         &   9.7     \\
\hline
RL PoWER Spline \cite{haocheng}                         & 300         & --         &   11.1 \\
\hline
GA + simulator \cite{glette}             & 60000       & 16.4       &   *13.0     \\
\hline
HyperNEAT + simulator [this paper]                     & 40000       & \textbf{25.4}       &   \textbf{14.5} \\
\hline
\end{tabular}
\end{center}
\label{resultsTable}
\end{table}

Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving,clune2011performance}. This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~(see \figref{servo_plot_111} for a representative example). This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation. This conclusion was not obvious after HyperNEAT failed to produce substantially coordinated, regular, effective gaits on this robot platform when evolving directly on the hardware~\cite{yos:clune}.

@@

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=fit_gen.eps, width=8cm}
% \caption[ ]{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s]. These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}
% \end{center}
% \end{figure}
\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{Performance plots. \textbf{Left: }Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s]. These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%. \edit{JBY: is it really standard deviation or standard error?}\textbf{Right: }Average frequencies of the gaits averaged over 20 runs. After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits. This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.\edit{JBY: fig might be easier to read if y-axis was log scale?}}

@@

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=servo_pos.eps, width=8cm}
% \caption[ ]{Top: A plot of servo positions for a gait produced in this study. Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware. Figure taken from Yosinski et al.~\cite{yos:clune}. The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
% \end{center}
% \end{figure}
%\figgp{servo_plot_111}{.6}{neat_110115_211410_00000_002_filt_zoom}{.6}{Top: A plot of servo positions for a gait produced in this study. Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware. Figure taken from Yosinski et al.~\cite{yos:clune}. The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}
\figp{servo_plot_111}{0.95}{Servo positions over time (for nine servos) for a gait produced in this study. The smoothness, symmetry, and regularities of the gait show that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}

### 2012-10-16 paper_sean/robot.tex
changed a few things, including table..and cut out some of my comments that sean has addressed
made every change as requested by Jeff, except for p from Wilcoxon rank sum test
JBY edits and comments.

% !TEX root =main.tex
%\figgp{quadratotWhiteBooties}{.45}{quadratot_simulator}{.45}{captioncaption}

We performed the experiments on the QuadraTot quadrupedal robot platform~\cite{yos:clune}.
It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline. %we def want a picture of the cool robot. I propose putting the picture of the robot on the left and the picture of the simulated robot on the right of the same figure. That also does a nice job of highlighting the theme of the paper (robot+sim) in one figure. Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page). 
QuadraTot has open source software and hardware\footnote{A list of all of the parts and all software, including the simulator, is available for free at http://quadratot.yosinski.com.} %JMC: This should really be a Creative Machines/Cornell web address. Can you ask Jason if he agrees and, if so, take the lead on making sure it gets made?
, including many parts that can be 3D printed. These traits enable other labs to perform research with it, providing a platform for comparing different gait-learning algorithms. 
To date there are results on the platform from nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}. We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote~\cite{yos:clune}.

@@

Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high~\cite{robotis}. During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process. As pointed out in Yosinski et al. 2011~\cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range. %JMC: Does requesting a value out of range really cause the servo to shut down? That seems like something we could easily fix (by never allowing an out-of-range value). Now that we do not allow values out of range, we still see lots of shutdowns...making me think that the problem was really torque. What do you think?
To minimize shutdowns, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively. We also implemented the Smart Cropping System from Shen et al. 2012~\cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque. A final method of reducing torque was to reduce the weight of the robot. Yosinski et al. 2011~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.

% % figure of the quadratot in its natural stance
% \begin{figure}
% \begin{center}
% \vspace{1cm}
% % just using a photo from Haocheng's paper! Should replace if needed be           
% \epsfig{file=quadratot_fig.eps, width=8cm}
% \caption[ ]{The QuadraTot robotic platform. It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world. The robot has 9 DOF, 2 per leg and a hip joint which connects the two symmetrical halves of the robot. The robot uses Dynamixel AX-18 and AX-12 servos for actuation. Figure taken from Shen et al. \cite{haocheng}.}
% \label{fig:quadratot}
% \end{center}
% \end{figure}
% %

%\figp{sixLegPositions}{.6}{Allowed and disallowed leg position angles. \edit{which are which? see JMC comment too} %JMC: Isn't this described in Haocheng's paper? I don't think we need to show this here, because it is not a new thing in our paper and is not central to our story. If we do keep it, instead of 3 sentences of text saying what the columns are, just put the headings ``Allowed'', ``End of Allowable Range'', and ``Disallowed''
%The figure shows an example of leg positions allowed and not allowed on the robot. The two left positions are allowed. The middle two positions are on the borderline. The right two positions are considered as ``fierce'' and are disallowed. Instead, these 
%Disallowed gaits are cropped to the middle two positions to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=gaitspace.eps, width=8cm}
% \caption[ ]{A map of possible leg positions. Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself. The sum of the inner and outer servo positions cannot drop below 730, the smart cropping border, marked by the red line. Figure taken from Shen et al. \cite{haocheng}.}
% \end{center}
% \end{figure}
%\figp{smartCrop}{.6}{A map of possible leg positions. Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself. The sum of the inner and outer servo positions cannot drop below 730, the smart cropping border, marked by the red line. Figure taken from Shen et al. \cite{haocheng}.}

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=wiimote.eps, width=8cm}
% \caption[ ]{The Wiimote used to track the motion of the robot. The Wiimote tracked an IR LED placed on top of the robot and sent the location data to a nearby computer which recorded the position of the robot with respect to time. Figure taken from Yosinski et al. \cite{yos:clune}.}
% \end{center}
% \end{figure}

%\figgp{wiiMote_crop.jpg}{.6}{robot_led_crop.jpg}{.6}{The Wiimote used to track the motion of ther robot. The Wiimote was used to track the IR LED placed on top of the robot and send the location data to a nearby computer which recorded the position of ther robot with respect to time. Figure taken from Yosinski et al. \cite{yos:clune}.}

>>>>

% !TEX root =main.tex
\subsubsection{Robot Hardware.}
%\figgp{quadratotWhiteBooties}{.45}{quadratot_simulator}{.45}{captioncaption}

We performed the experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties}).
It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline. %we def want a picture of the cool robot. I propose putting the picture of the robot on the left and the picture of the simulated robot on the right of the same figure. That also does a nice job of highlighting the theme of the paper (robot+sim) in one figure. Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page). 
The QuadraTot hardware designs and software for this project are open source\footnote{A list of all of the parts and all software, including the simulator, is available for free at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts. These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms. 
To date there are results on the platform from nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}. We identified the robot's position using an infrared LED on the robot which was tracked using a Wiimote~\cite{yos:clune}.

@@

Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high~\cite{robotis}. During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process. As pointed out in Yosinski et al. 2011~\cite{yos:clune}, \edit{is it better to say ``As pointed out in a previous study \cite{yos:clune}, gaits...''} gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range. %JMC: Does requesting a value out of range really cause the servo to shut down? That seems like something we could easily fix (by never allowing an out-of-range value). Now that we do not allow values out of range, we still see lots of shutdowns...making me think that the problem was really torque. What do you think?
To prevent collisions between different pieces of the robot's body, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively. We also implemented the Smart Cropping System from Shen et al. 2012~\cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque. A final method of reducing torque was to reduce the weight of the robot. Yosinski et al. 2011~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.

### 2012-10-16 paper_sean/simulator.tex
JBY edits and comments.

% !TEX root =main.tex
The simulator was the same as Glette et al. 2012 \cite{glette}. C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}). In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.

>>>>

% !TEX root =main.tex

\subsubsection{Simulator.}
The simulator was the same as Glette et al. 2012 \cite{glette}. C++ code for the the Nvidia PhysX physics engine\footnotemark[\value{footnote}]
 approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}, right). In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.
